Volunteers in the Clouds: A Web Based Architecture to Harness Volunteer Resources for Low Cost Distributed Evolutionary Computation

### 2016-01-11 special-issue-ci/volunteer-clouds-ieee-ci-si.tex
Post-final changes
Under the request of the editor.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-7-31.png}
\caption{Duration of experiments vs. number of different IPs (nodes)
  participating in it, with averages and standard deviation shown as
  red dots; in the case there is a single red dot, there was a single
  experiment in which many computers participated (for instance, 16
  computers in the experiment in the far left or 29 in the middle
  one). 
Shades of gray indicate how many experiments included that many unique IPs,
so lighter shade for a column of dots indicates that a particular number
of computers happened less frequently, while darker shadow means more frequency. 
From left to right, experiments 4/4, 4/24 and 7/31.}
\label{fig:duration}
\end{figure}
% ../data/time-vs-IPs-openshift.R
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-7-31.png}
\caption{Duration of experiments vs. rank, with $y$ axis in a
  logarithmic scale. Dot color (or gray value) is related to the number of IPs
  participating in the experiment. From left to right, experiments
  4/4, 4/24 and 7/31.} 
\label{fig:zipf:os}
% computed via ../data/time-vs-IPs-openshift.R
\end{figure}
%
We will have to analyze experimental data a bit further to find out why
this happens and also if there are some patterns in the three sets of
experiments. An interesting question to ask, for instance, is if
by adding more computers makes the experiment take less. In fact, as
shown in Figure \ref{fig:duration}, the {\em addition} of more computers does
not seem to contribute to decreasing the time needed to finish the
experiment. However, the cause-effect relationship is not clear at
all. It might be the opposite: since experiments take longer to finish
and might in fact be abandoned with no one contributing for some time,
that increases the probability of someone new joining them. In fact,
with experiments taking a few seconds and due to the way the
experiments are announced, it is quite difficult that several
volunteers join in in such a short period of time, even more if we take
into account that volunteers are not {\em carried over} from previous
experiments. This implies that it would be convenient to use a problem
of a bigger size to check this hypothesis as we have done in the next
experiment. 
%  however, at this point we
% have not found this convenient since there are several other issues
% that have to be solved, as it will be shown next. %Creo que este párrafo
%saldría sobrando ahora que hicmios un experimento más grande

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{puts-openshift-4-4.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-4-24.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-7-31.png}
\caption{Number of {\tt PUT}s per unique IP and fit to a Generalized
  Extreme Value distribution (in a lighter shade of gray). From left to right, experiments
  4/4, 4/24 and 7/31.} 
% Plotted with ../data/plot-zipf-openshift.R
\label{fig:puts:os}
\end{figure}
%
It is also interesting to check the distribution of the experiment
duration, shown in Figure \ref{fig:zipf:os} and which roughly follows
a Zipf's law, with similar distribution along all three runs. The 4/24
run is the most complete and shows a S-shape, which implies an
accumulation of experiments taking similar time and around 100
seconds. The most interesting part is the {\em tail}, which shows how
many experiments took a desirable amount of time, of the order of
10 seconds, and which appears in all three graphs. As it can be seen,
it sharply drops implying there are 
just a few of them, and with diminishing probability as time
decreases. This exponential distribution also appears in the
distribution of HTTP {\tt PUT}s, equivalent to the number of
generations divided by 100, contributed by every user, which is shown
in Figure \ref{fig:puts:os}. These results show a Zipf-like behavior,
so we have fitted it to the Generalized Extreme Value distribution,
with the resulting parameters shown in Table \ref{tab:puts:os}.
%
\begin{table*}
\caption{Summary of fit to Generalized Extreme Value distribution of
  the number of {\tt PUT}s per unique IP. \label{tab:puts:os}}
\begin{center}
\begin{tabular}{l|ccc}
\hline
Date  & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
4/4 &  8.541 $\pm$ 1.0926  &    12.442 $\pm$ 1.7302 &  1.388 $\pm$
0.1377 \\
4/24 & 6.148 $\pm$ 0.3782 & 7.354 $\pm$ 0.5105 & 1.090 $\pm$  0.0697  \\
7/31 & 11.645 $\pm$ 1.475 & 16.365 $\pm$ 2.201 &  1.265 $\pm$ 0.132   \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
This distribution was originally proposed to fit extreme values
\cite{resnick2013extreme} and contains, as a special case, the inverse
Weibull distribution which was fitted to volunteer computing
frameworks such as SETI@home \cite{javadi2009mining}. We obviously do
not pretend to compare our framework in scale or complexity with it, but
to point out that the behavior of volunteer computing nodes follows a
certain pattern, found in SETI@home, and which also appears in our framework. This
distribution is governed by three parameters, the usual location $\mu$
which is related to where it has its {\em center} and a scale $\sigma$,
related to the size, but also a third shape $\xi$ parameter that is
related to its skewness, that is, how skewed it is around the central
location. Positive parameters indicate that the distribution {\em
  leans} towards the origin, and negative ones towards the other extreme
value. In this case, Table \ref{tab:puts:os} shows $\xi$ values
greater than one and between one and 1.4, which indicates that the
three experiments share this origin-leaning pattern, with many users
donating a few cycles and just a few donating extreme values. Random
distributions with these parameters have been plotted in red in
Figure \ref{fig:puts:os}, indicating that the fit is good enough. The
predictive value of these fits, however, is limited, over all taking
into account the low correlation between successive events shown
above.

@@

\begin{table*}[!htb]
\caption{Summary of {\sf NodIO-W$^2$} and comparison with the previous
  experiments, which have been aggregated to compute central measures. \label{tab:summary:ww}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \#{\tt
  PUT}s & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline
{\sf NodIO} & 5 & 29 & 123 & 14 & 37.43\% & 3.40\% & 0.10 \\
{\sf NodIO-W$^2$} & 4  & 16 & 7.36 & 40 & 89\% & 36.90\% & 0.4336061 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.9\linewidth]{ips-time-ww.png}
\caption{Time employed in every experiment vs. number of IPs (in
  abscissas). The black dot and line show the average and standard
  deviation, the shades of gray the number of experiments with the same
  number of IPs. } 
\label{fig:ipstime:w2}
% ../data/time-vs-IPs.R 
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.49\linewidth]{gev-fit-ww.png}
\includegraphics[width=0.49\linewidth]{weibull-fit-ww.png}
\caption{Ranked number of {\tt PUT}s per worker and GEV fit (in light color, right) 
and Weibull (in lighter shade, left) for the {\sf NodIO-W$^2$} model.}  
\label{fig:gev:w2}
% File ../data/ips-puts.R
\end{figure}
%
As we did in the previous subsection, we have also fitted the number
of HTTP {\tt PUT}s per worker to a  Generalized Extreme Value (GEV) function, with the result shown
in Table \ref{tab:puts:ww}. Comparing this table with what we obtained
in the previous experiment, Table \ref{tab:puts:os}, we find higher values
for the location $\mu$ and scale $\sigma$ parameters, accounting for
the higher number of contributions per volunteer that has been
previously observed. However, the shape $\xi$ parameter is
substantially similar, with a {\em bell} shape leaning towards the
origin, indicating a pattern of a few users with many contributions
and many with few contributions. If we plot the model and the data
side by side, as shown in Figure \ref{fig:gev:w2} we see that the fit
is not so tight as in the previous case, with the model overestimating
the number of individuals with a high number contributions. This is
probably due to the fact that in the previous case, the user needed to
voluntarily reload the page to contribute the most generations, with a
{\em phase change} between the people that did so and the people that
just ran the experiment once until completion. In this case, however, the
user just needs to let it run, without that phase change and thus we obtain a
more {\em egalitarian} distribution of contributions, which rather
corresponds to a Weibull function, as shown in Figure
\ref{fig:ipstime:w2} (right hand side). The fitted parameters for this
distribution, shown in Table \ref{tab:puts:ww}, show a shape parameter
equal to 0.69, a value that is remarkably similar to the 0.5 value
found for time devoted to games in Chambers et
al. \cite{chambers2005measurement}. It is important to note that, in
this case, the number of PUTs made per client does not correspond to a
uniform amount of computation, since 100 generations with a random
population belong, in each case, to different number of
operations. In other words, the same number of PUTs will correspond to
different number of operations in each case. However, since clients
are also different and take a different time in each case, it is
difficult to ascertain how this might have an influence in the
statistical distribution.

@@

Finally, the implementation needs some refinement in terms of
programming and also ease of use. Tools such as Yeoman might be used % ¿añadir una cita o la URL http://yeoman.io/ a pie de página ?
to create a generator in which the user just has to create a fitness
function, with the rest of the framework wrapped around
automatically. All this, as well as the data used for this paper and the paper
itself, has been published with a free license in GitHub at
\url{https://github.com/JJ/modeling-volunteer-computing}.

>>>>

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-7-31.png}
\caption{Duration of experiments vs. number of different IPs (nodes)
  participating in it, with averages and standard deviation shown as
  red dots; in the case there is a single dot, there was a single
  experiment in which many computers participated (for instance, 16
  computers in the experiment in the far left or 29 in the middle
  one). 
Shades of gray indicate how many experiments included that many unique IPs,
so lighter shade for a column of dots indicates that a particular number
of computers happened less frequently, while darker shade or closer to
redmeans more frequency. 
From left to right, experiments 4/4, 4/24 and 7/31.}
\label{fig:duration}
\end{figure}
% ../data/time-vs-IPs-openshift.R
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-7-31.png}
\caption{Duration of experiments vs. rank, with $y$ axis in a
  logarithmic scale. Dot color (or gray value) is related to the number of IPs
  participating in the experiment. From left to right, experiments
  4/4, 4/24 and 7/31.} 
\label{fig:zipf:os}
% computed via ../data/time-vs-IPs-openshift.R
\end{figure}
%
We will have to analyze experimental data a bit further to find out why
this happens and also if there are some patterns in the three sets of
experiments. An interesting question to ask, for instance, is if
by adding more computers makes the experiment take less. In fact, as
shown in Figure \ref{fig:duration}, the {\em addition} of more computers does
not seem to contribute to decreasing the time needed to finish the
experiment. However, the cause-effect relationship is not clear at
all. It might be the opposite: since experiments take longer to finish
and might in fact be abandoned with no one contributing for some time,
that increases the probability of someone new joining them. In fact,
with experiments taking a few seconds and due to the way the
experiments are announced, it is quite difficult that several
volunteers join in in such a short period of time, even more if we take
into account that volunteers are not {\em carried over} from previous
experiments. This implies that it would be convenient to use a problem
of a bigger size to check this hypothesis as we have done in the next
experiment. 
%  however, at this point we
% have not found this convenient since there are several other issues
% that have to be solved, as it will be shown next. %Creo que este párrafo
%saldría sobrando ahora que hicmios un experimento más grande

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{puts-openshift-4-4.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-4-24.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-7-31.png}
\caption{Number of {\tt PUT}s per unique IP and fit to a Generalized
  Extreme Value distribution (in a lighter shade or blue). From left to right, experiments
  4/4, 4/24 and 7/31.} 
% Plotted with ../data/plot-zipf-openshift.R
\label{fig:puts:os}
\end{figure}
%
It is also interesting to check the distribution of the experiment
duration, shown in Figure \ref{fig:zipf:os} and which roughly follows
a Zipf's law, with similar distribution along all three runs. The 4/24
run is the most complete and shows a S-shape, which implies an
accumulation of experiments taking similar time and around 100
seconds. The most interesting part is the {\em tail}, which shows how
many experiments took a desirable amount of time, of the order of
10 seconds, and which appears in all three graphs. As it can be seen,
it sharply drops implying there are 
just a few of them, and with diminishing probability as time
decreases. This exponential distribution also appears in the
distribution of HTTP {\tt PUT}s, equivalent to the number of
generations divided by 100, contributed by every user, which is shown
in Figure \ref{fig:puts:os}. These results show a Zipf-like behavior,
so we have fitted it to the Generalized Extreme Value distribution,
with the resulting parameters shown in Table \ref{tab:puts:os}.
%
\begin{table*}
\caption{Summary of fit to Generalized Extreme Value distribution of
  the number of {\tt PUT}s per unique IP. \label{tab:puts:os}}
\begin{center}
\begin{tabular}{l|ccc}
\hline
Date  & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
4/4 &  8.541 $\pm$ 1.0926  &    12.442 $\pm$ 1.7302 &  1.388 $\pm$
0.1377 \\
4/24 & 6.148 $\pm$ 0.3782 & 7.354 $\pm$ 0.5105 & 1.090 $\pm$  0.0697  \\
7/31 & 11.645 $\pm$ 1.475 & 16.365 $\pm$ 2.201 &  1.265 $\pm$ 0.132   \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
This distribution was originally proposed to fit extreme values
\cite{resnick2013extreme} and contains, as a special case, the inverse
Weibull distribution which was fitted to volunteer computing
frameworks such as SETI@home \cite{javadi2009mining}. We obviously do
not pretend to compare our framework in scale or complexity with it, but
to point out that the behavior of volunteer computing nodes follows a
certain pattern, found in SETI@home, and which also appears in our framework. This
distribution is governed by three parameters, the usual location $\mu$
which is related to where it has its {\em center} and a scale $\sigma$,
related to the size, but also a third shape $\xi$ parameter that is
related to its skewness, that is, how skewed it is around the central
location. Positive parameters indicate that the distribution {\em
  leans} towards the origin, and negative ones towards the other extreme
value. In this case, Table \ref{tab:puts:os} shows $\xi$ values
greater than one and between one and 1.4, which indicates that the
three experiments share this origin-leaning pattern, with many users
donating a few cycles and just a few donating extreme values. Random
distributions with these parameters have been plotted in red in
Figure \ref{fig:puts:os}, indicating that the fit is good enough. The
predictive value of these fits, however, is limited, over all taking
into account the low correlation between successive events shown
above.

@@

\begin{table*}[!htb]
\caption{Summary of {\sf NodIO-W$^2$} and comparison with the previous
  experiments, which have been aggregated to compute central measures. \label{tab:summary:ww}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \#{\tt
  PUT}s & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline
{\sf NodIO} & 5 & 29 & 123 & 14 & 37.43\% & 3.40\% & 0.10 \\
{\sf NodIO-W$^2$} & 4  & 16 & 7.36 & 40 & 89\% & 36.90\% & 0.4336061 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.9\linewidth]{ips-time-ww.png}
\caption{Time employed in every experiment vs. number of IPs (in
  abscissas). The black dot and line show the average and standard
  deviation, the blue to red shades the number of experiments with the same
  number of IPs. } 
\label{fig:ipstime:w2}
% ../data/time-vs-IPs.R 
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.49\linewidth]{gev-fit-ww.png}
\includegraphics[width=0.49\linewidth]{weibull-fit-ww.png}
\caption{Ranked number of {\tt PUT}s per worker and GEV fit (in light
  or blue color, left) 
and Weibull (in lighter or blue shade, right) for the {\sf NodIO-W$^2$} model.}  
\label{fig:gev:w2}
% File ../data/ips-puts.R
\end{figure}
%
As we did in the previous subsection, we have also fitted the number
of HTTP {\tt PUT}s per worker to a  Generalized Extreme Value (GEV) function, with the result shown
in Table \ref{tab:puts:ww}. Comparing this table with what we obtained
in the previous experiment, Table \ref{tab:puts:os}, we find higher values
for the location $\mu$ and scale $\sigma$ parameters, accounting for
the higher number of contributions per volunteer that has been
previously observed. However, the shape $\xi$ parameter is
substantially similar, with a {\em bell} shape leaning towards the
origin, indicating a pattern of a few users with many contributions
and many with few contributions. If we plot the model and the data
side by side, as shown in Figure \ref{fig:gev:w2} we see that the fit
is not so tight as in the previous case, with the model overestimating
the number of individuals with a high number contributions. This is
probably due to the fact that in the previous case, the user needed to
voluntarily reload the page to contribute the most generations, with a
{\em phase change} between the people that did so and the people that
just ran the experiment once until completion. In this case, however, the
user just needs to let it run, without that phase change and thus we obtain a
more {\em egalitarian} distribution of contributions, which rather
corresponds to a Weibull function, as shown in Figure
\ref{fig:ipstime:w2} (right hand side). The fitted parameters for this
distribution, shown in Table \ref{tab:puts:ww}, show a shape parameter
equal to 0.69, a value that is remarkably similar to the 0.5 value
found for time devoted to games in Chambers et
al. \cite{chambers2005measurement}. It is important to note that, in
this case, the number of PUTs made per client does not correspond to a
uniform amount of computation, since 100 generations with a random
population belong, in each case, to different number of
operations. In other words, the same number of PUTs will correspond to
different number of operations in each case. However, since clients
are also different and take a different time in each case, it is
difficult to ascertain how this might have an influence in the
statistical distribution.

@@

Finally, the implementation needs some refinement in terms of
programming and also ease of use. Tools such as Yeoman might be used % ¿añadir una cita o la URL http://yeoman.io/ a pie de página ?
to create a generator in which the user just has to create a fitness
function, with the rest of the framework wrapped around
automatically.
%  All this, as well as the data used for this paper and the paper
% itself, has been published with a free license in GitHub at
% \url{https://github.com/JJ/modeling-volunteer-computing}.
